\documentclass[11pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{enumitem}

\title{MedFailBench: A Clinician-Built Open-Source Benchmark for\\Medical AI Safety Boundary Inspection}

\author[1]{Goktug Ozkan, MD}
\affil[1]{Department of Internal Medicine, Kutahya Emet Dr. Fazil Dogan State Hospital, Kutahya, Turkiye}
\date{Draft v0.2}

\begin{document}
\maketitle

\begin{abstract}
Most medical AI benchmarks measure whether a model knows the correct answer.
MedFailBench asks a different question: which safety boundary failed?
We present a clinician-built synthetic benchmark and failure atlas that labels
medical AI errors by severity (1--5) and safety gate type (missed urgent
escalation, unsafe remote dosing, unsafe discharge reassurance, evidence
fabrication, unsafe protocol execution, source support gap). The current public
release contains 44 clinician-reviewed synthetic cases with severity annotations,
a live HuggingFace leaderboard preview, a safety gate taxonomy, a clinical
severity rubric, and the first real model response evaluation covering 3 models
across 5 hard prompts. No patient data, no clinical validation claims, no model
rankings. MedFailBench is released under Apache-2.0 and CC-BY-4.0.
\end{abstract}

\section{Introduction}

Large language models (LLMs) are increasingly proposed for clinical settings.
General benchmarks such as MMLU~\cite{hendrycks2021mmlu} measure knowledge
retrieval, but medical safety failures often manifest as wording, escalation
boundary, source support, and clinical framing errors that exam accuracy metrics
do not capture~\cite{wornow2025healthbench}.

Existing medical AI safety evaluations have examined specific failure modes
such as fabricated evidence~\cite{pal2024medhelm}, unsafe dosing
advice~\cite{levkovich2024dosing}, and patient-facing safety
failures~\cite{shaib2024llmclinicalsafety}. Single-model assessments of
GPT-4~\cite{nori2023gpt4medical,lee2023benefitslimitsgpt4} and Med-PaLM
2~\cite{singhal2023medpalm2} in clinical tasks have been published, but
most open benchmarks lack a structured clinician-reviewed severity layer that
separates wording problems from safety-critical framing errors.

The Turkish National AI Strategy 2021--2025 identifies healthcare AI as a
priority area, and the 2026--2030 AI Vision includes a national LLM (Bilge)
and a health AI conformity assessment framework. MedFailBench is designed as a
public safety review layer that can inform these national efforts while
remaining independent of any institutional endorsement.

We introduce MedFailBench, an open-source synthetic benchmark that labels
medical AI failure patterns by clinician-reviewed severity and safety gate
type. The benchmark is designed to make failure patterns inspectable without
requiring patient data or institutional approval.

\section{Methods}

\subsection{Synthetic Case Construction}

Each case consists of:
\begin{enumerate}[noitemsep]
  \item A synthetic clinical prompt with variables intentionally left open.
  \item The output generated by a medical AI system (model or synthetic).
  \item A clinician-authored description of the boundary failure.
  \item A severity score (1--5) and safety gate label.
\end{enumerate}

Cases are constructed by the lead clinician using anonymized clinical
reasoning patterns drawn from general internal medicine practice. Each
prompt is designed to test a specific boundary: urgency escalation,
medication safety, discharge reassurance, evidence support, or protocol
execution. Prompts include open variables (e.g., vital signs, lab values)
that a safe system must recognize as missing before clinical action.

No patient records, real clinical notes, or identifiable data are used.
All cases were reviewed by Goktug Ozkan, MD (Internal Medicine).

\subsection{Severity Rubric}

Severity levels describe the clinical risk represented by one case or one
reviewed output, not a global model safety claim.

\begin{table}[h]
\centering
\begin{tabular}{p{1.5cm} p{3cm} p{6cm}}
\toprule
Level & Label & Meaning \\
\midrule
1 & Wording issue & Awkward or incomplete wording with low clinical risk. \\
2 & Missing caution & Answer misses a caution or boundary a clinician would expect. \\
3 & Clinical ambiguity risk & Key variables or escalation boundaries missing; could mislead. \\
4 & Safety-critical miss & Could delay urgent evaluation or normalize unsafe action. \\
5 & High-risk unsafe framing & Gives or implies unsafe action, false reassurance, or emergency deferral. \\
\bottomrule
\end{tabular}
\caption{Clinical severity rubric v0.2.}
\end{table}

Distribution in the current synthetic set: severity 3 = 7 cases, severity 4 = 14 cases,
severity 5 = 23 cases.

\subsection{Safety Gate Taxonomy}

A safety gate names the primary review reason that connects a model failure to
the clinical risk a reviewer needs to inspect~\ref{tab:gates}.

\begin{table}[h]
\centering
\begin{tabular}{p{5cm} p{6cm}}
\toprule
Gate & Definition \\
\midrule
missed\_urgent\_escalation & Answer fails to preserve urgent evaluation. \\
unsafe\_remote\_dosing & Answer implies medication change without missing variables. \\
unsafe\_discharge\_reassurance & Reassurance despite unresolved red flags. \\
evidence\_fabrication\_or\_overclaim & Overstates evidence or invents support. \\
unsafe\_protocol\_execution\_detail & Turns general guidance into operational steps. \\
source\_support\_gap & Source support problem needing explicit checking. \\
\bottomrule
\end{tabular}
\caption{Safety gate taxonomy v0.2.\label{tab:gates}}
\end{table}

\subsection{Real Model Response Evaluation}

We evaluated 3 open-source LLMs (DeepSeek V4 Flash, Qwen 2.5 7B Instruct,
Llama 3.3 70B Instruct) on 5 hard synthetic prompts from the MedFailBench set.
Each output was reviewed using the severity rubric and safety gate taxonomy.
Results were documented as a weekly evaluation preview and made available in
the repository.

\subsection{Repository Structure}

The MedFailBench repository (\url{https://github.com/goktugozkanmd/medical-ai-failure-atlas})
is organized for inspectability and contribution:
\begin{enumerate}[noitemsep]
  \item \texttt{data/} -- synthetic case files in structured JSON format with
  severity and gate annotations.
  \item \texttt{leaderboard/} -- HuggingFace Space source for the live severity
  leaderboard preview.
  \item \texttt{model\_runs/} -- weekly model response evaluation results and
  scoring scripts.
  \item \texttt{preprint/} -- LaTeX source for this document with \texttt{references.bib}.
  \item \texttt{scripts/} -- benchmarking, scoring, and validation scripts.
  \item \texttt{docs/} -- collaboration brief, roadmap, outreach drafts, and
  contribution templates.
\end{enumerate}
All synthetic cases are licensed under CC-BY-4.0; code is Apache-2.0.

\section{Results}

The current public release includes:
\begin{enumerate}[noitemsep]
  \item A synthetic-only case set with 44 clinician-reviewed entries.
  \item A safety gate taxonomy with 6 defined gate types.
  \item A clinical severity rubric with 5 levels.
  \item A live HuggingFace leaderboard preview with model submission form.
  \item A first weekly model response evaluation covering 3 models x 5 prompts.
  \item A collaboration brief and contribution guide with issue templates.
\end{enumerate}

All cases are severity-annotated. The distribution is weighted toward higher
severity: 23 of 44 cases (52\%) are level 5 (high-risk unsafe framing).
The most common safety gate is missed urgent escalation (16/44, 36\%),
consistent with the observation that LLMs may produce fluent text without
preserving escalation boundaries.

\begin{table}[h]
\centering
\begin{tabular}{l r r r r r r}
\toprule
Model & Safety & Acc. & Source & Refusal & Clin. & Auto final \\
& mean & mean & trans. & approp. & ground. & labels \\
\midrule
DeepSeek V4 Flash & 1.8 & 2.2 & 4.0 & 4.2 & 3.6 & 4 unsafe; 1 caution \\
Qwen 2.5 7B Instr. & 1.8 & 2.2 & 4.0 & 4.2 & 3.6 & 4 unsafe; 1 caution \\
Llama 3.3 70B Instr. & 1.6 & 2.2 & 4.0 & 4.2 & 4.0 & 4 unsafe; 1 caution \\
\bottomrule
\end{tabular}
\caption{Automated rule-based evaluation of 3 models across 5 hard synthetic prompts (H001--H005). Scores are 1--5 (5 best). This is a preliminary automation check; clinician review is required before any public claim.}
\end{table}

\section{Discussion}

MedFailBench provides a clinician-reviewed severity and safety gate layer
that existing medical AI benchmarks do not offer. While benchmarks such as
MMLU~\cite{hendrycks2021mmlu} and HealthBench~\cite{wornow2025healthbench}
report aggregate knowledge accuracy, they do not distinguish between a
fluent but unsafe response and a cautious refusal. MedFailBench shifts the
question from \textit{does the model know the answer?} to \textit{what safety
boundary failed and how dangerous is it?}

The preliminary automated evaluation demonstrates that all three tested
models produce safety-related failures on the same hard prompts, with missed
urgent escalation as the dominant failure mode. This consistency across
model sizes (7B to 70B parameters) and architectures suggests that medical
safety boundary failures are not limited to small models but reflect a
deeper pattern in how LLMs handle clinical ambiguity. Full clinician review
is required to validate these automated findings and to produce a detailed
failure profile.

\subsection{Relation to Other Benchmarks}

MedFailBench is complementary to existing medical benchmarks. MMLU and
MedMCQA measure factual knowledge; MedHELM~\cite{pal2024medhelm} evaluates
multiple dimensions including calibration and fairness; HealthBench assesses
clinical knowledge across specialties. None of these provide a structured
clinician-reviewed severity layer that separates wording problems from
safety-critical framing errors.

Specialized evaluations of GPT-4~\cite{nori2023gpt4medical} and Med-PaLM
2~\cite{singhal2023medpalm2} have examined long-form clinical reasoning, but
their findings are not packaged as open, reusable benchmark artifacts with
inspectable failure labels. MedFailBench fills this gap by making each
failure pattern inspectable, extensible, and community-reviewable.

For Turkish-language medical AI evaluation, TR-MMMU~\cite{tsai2025trmmmu}
and recent literature surveys~\cite{kuly2024turkhangi} provide context for
the national dimension of this work. MedFailBench extends these efforts by
adding a safety layer that is independent of language-specific knowledge
benchmarks.

\subsection{Relation to Turkish AI Strategy}

The Turkish 2026--2030 AI Vision identifies health AI conformity assessment
and the Bilge national LLM as strategic priorities. MedFailBench can function
as an independent public safety review surface for these efforts, providing
transparent failure inspection without patient data access. We position
MedFailBench as a community resource that national initiatives can build on
rather than a certified assessment tool.

\subsection{Future Work}

Several directions are planned for v0.2.1 and beyond. First, a permanent DOI
via Zenodo will anchor academic citation. Second, automated weekly evaluation
runs will track model behavior changes over time. Third, a second clinician
reviewer will provide inter-rater reliability data, addressing the main
limitation of single-clinician annotation. Fourth, the synthetic case set
will be expanded toward 100 cases with broader specialty coverage. Fifth,
a lightweight submission API will allow model providers to run their systems
against MedFailBench and receive automated safety reports. Community
contributions through synthetic cases, safety gate objections, and wording
review are welcome and coordinated through GitHub Discussions.

\section{Limitations}

\begin{enumerate}[noitemsep]
  \item Synthetic cases only; no real patient data and no demonstrated
  generalizability to real clinical workflows.
  \item Single-clinician review with no inter-rater reliability measurement.
  \item Small case count (44 entries) with skewed severity distribution.
  \item Not a clinical validation study, model ranking, or deployment safety
  certification.
  \item The weekly model response evaluation is a preview, not a longitudinal
  or controlled evaluation.
  \item The automated rule-based scoring system requires validation against
  clinician review before any performance claims can be made.
\end{enumerate}

\section{Conclusion}

Clinician-led severity annotation and safety gate taxonomy can help make
medical AI failure modes easier to inspect, discuss, and improve in open-source
settings. The MedFailBench approach shifts focus from model accuracy to safety
boundary analysis, providing a structured surface for community review that
existing accuracy-only benchmarks do not support. We invite clinicians,
researchers, and safety reviewers to contribute cases, gate objections, and
wording reviews through the public repository and GitHub Discussions.
MedFailBench is released as an open resource under Apache-2.0 and CC-BY-4.0.

\nocite{*}
\bibliographystyle{plain}
\bibliography{references}

\section*{Data and Code Availability}
MedFailBench v0.2.0 is released under Apache-2.0 and CC-BY-4.0.
All source code, synthetic cases, evaluation data, and the live leaderboard
preview are available at \url{https://github.com/goktugozkanmd/medical-ai-failure-atlas}
and \url{https://huggingface.co/spaces/goktugozkanmd/medical-ai-failure-atlas}.
A Zenodo DOI will be assigned at v0.2.1.

\end{document}